---
title: "Citizens prefer mandatory over voluntary biodiversity financing even under favorable conditions"
author:
  - name: Nils Droste
    affiliations: Department of Food and Resource Economics, University of Copenhagen
    email: nd@ifro.ku.dk
  - name: Jens Marl Christiansen
    affiliations: Department of People and Technology, Roskilde University
    email: jmchristiansen@ruc.dk
  - name: Yohei Mitani
    affiliations: Graduate School of Global Environmental Studies, Kyoto University
    email: mitani.yohei.7w@kyoto-u.ac.jp
date: today
abstract: |
  Global biodiversity governance has, over three decades, shifted substantially toward voluntary, market-based, and non-state financing instruments. While academic and policy evidence on the supply side of this shift has been growing, the demand side is largely missing from the debate: do citizens actually support voluntary biodiversity financing schemes over mandatory ones? We test this in Sweden, which provides the most structurally favorable context for voluntary instrument support globally, combining high social trust, strong environmental identity, and among-the-highest per-capita conservation spending. If citizen demand for voluntary instruments fails here, it fails at its own best case. Using a preregistered discrete choice experiment ($n = 2{,}101$) combined with an incentivised donation experiment, we find that Swedish citizens consistently prefer mandatory biodiversity financing instruments over voluntary alternatives. Corporate offsetting carries the highest mean utility and public monitoring commands a willingness-to-pay premium of approximately SEK 1,600 per year. Citizens who care most about biodiversity outcomes — and who personally contribute to conservation — prefer mandatory instruments most strongly, ruling out strategic free-riding. Since we cannot expect voluntary contributions from less engaged citizens either, voluntary instruments lack a natural constituency. We recommend mandatory instruments for biodiversity financing in Western countries with sufficiently high state trust, and theorize that voluntary instrument preference elsewhere reflects constrained institutional choice rather than genuine endorsement of voluntary governance.
keywords:
  - biodiversity financing
  - discrete choice experiment
  - voluntary governance
  - most-likely case design
  - Kunming-Montreal Global Biodiversity Framework
format:
  pdf:
    number-sections: false
    fontsize: 12pt
    geometry: margin=2.5cm
    keep-tex: false
  docx:
    number-sections: false
bibliography: references.bib
execute:
  eval: false
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| eval: true
library(tidyverse)
library(knitr)
library(kableExtra)
library(patchwork)
library(here)
library(readxl)
library(sandwich)
library(lmtest)
```

```{r}
#| label: load-data
#| eval: true

data_path   <- here::here("data", "Full launch database", "4178_excel_databas.xlsx")
design_path <- here::here("design", "Trial 3 - factorial grouped, svensk.xlsx")

full_data  <- read_excel(data_path, sheet = 2)
price_map  <- c("0" = 492, "1" = 2460, "2" = 4920, "3" = 7380)
design_raw <- read_excel(design_path, sheet = 1)
names(design_raw) <- c(
  "cs","block","a_source","a_src_txt","a_land","a_land_txt",
  "a_monitor","a_mon_txt","a_price","a_price_txt",
  "b_source","b_src_txt","b_land","b_land_txt",
  "b_monitor","b_mon_txt","b_price","b_price_txt"
)
design <- design_raw |>
  select(cs, block, a_source, a_land, a_monitor, a_price,
         b_source, b_land, b_monitor, b_price) |>
  group_by(block) |> mutate(task = row_number()) |> ungroup() |>
  mutate(a_price_sek = price_map[as.character(a_price)],
         b_price_sek = price_map[as.character(b_price)])

choices_ord1 <- full_data |> filter(ordning == 1) |>
  select(id, block, choice1:choice8) |>
  pivot_longer(choice1:choice8, names_to = "col", values_to = "choice") |>
  mutate(task = as.integer(substr(col, 7, 7)))
choices_ord2 <- full_data |> filter(ordning == 2) |>
  select(id, block, choice12, choice22, choice32, choice42,
         choice52, choice62, choice72, choice82) |>
  pivot_longer(-c(id, block), names_to = "col", values_to = "choice") |>
  mutate(task = as.integer(substr(col, 7, 7)))

resp <- bind_rows(choices_ord1, choices_ord2) |>
  filter(!is.na(choice)) |>
  left_join(design, by = c("block", "task")) |>
  mutate(choice_num = case_when(choice == "a" ~ 1L, choice == "b" ~ 2L, TRUE ~ 3L))

resp_level <- resp |>
  mutate(
    a_vol = a_source %in% c(1L, 2L), b_vol = b_source %in% c(1L, 2L),
    a_man = a_source %in% c(0L, 3L), b_man = b_source %in% c(0L, 3L),
    mixed_task = (a_vol & b_man) | (a_man & b_vol),
    chose_vol  = case_when(
      !mixed_task | choice == "3" ~ NA_real_,
      (a_vol & choice == "a") | (b_vol & choice == "b") ~ 1, TRUE ~ 0
    )
  ) |>
  filter(mixed_task, !is.na(chose_vol)) |>
  group_by(id) |>
  summarise(voluntary_pref = mean(chose_vol), n_mixed = n(), .groups = "drop")

covars <- full_data |>
  transmute(
    id,
    donation    = yourdonation,
    gave        = as.integer(yourdonation > 0),
    env_import  = q11,
    income      = q8,
    gender      = q10,
    trust_state = q2_1,
    trust_ngo   = q2_3,
    policy_cq   = q14,
    bio_knowl   = q12
  )

anal_data <- resp_level |>
  left_join(covars, by = "id") |>
  filter(!is.na(donation), !is.na(env_import), !is.na(income), !is.na(gender),
         !is.na(trust_state), !is.na(trust_ngo), !is.na(policy_cq), !is.na(bio_knowl)) |>
  mutate(
    env_med     = median(env_import),
    env_hi      = env_import >= env_med,
    donated     = gave == 1L,
    env_z       = as.numeric(scale(env_import)),
    donation_z  = as.numeric(scale(donation)),
    gave_binary = gave,
    income_mid  = as.integer(income %in% 2:4),
    income_high = as.integer(income %in% 5:7),
    female      = as.integer(gender == 2),
    env_t       = ntile(env_import, 3),
    env_lab     = factor(env_t, labels = c("Low env. att.", "Medium", "High env. att."))
  )
```

# Introduction

The Kunming-Montreal Global Biodiversity Framework (GBF) commits governments to mobilising at least USD 200 billion per year for nature by 2030, relying heavily on voluntary, market-based, and non-state financing instruments alongside public funding [@CBD2022]. Private donations to conservation organisations, consumer certification schemes, and voluntary corporate offsetting have been promoted as flexible, decentralised alternatives to mandatory state regulation. The supply side of this voluntary governance architecture has attracted considerable academic and policy attention: whether voluntary biodiversity markets can scale, whether certification standards hold, and whether non-state actors can credibly commit [@vanderVen2022; @Maron2025].

The demand side has received far less attention. Does citizen demand support voluntary instruments as the governance architecture does? For voluntary biodiversity financing to function, two conditions must hold: citizens who care about biodiversity outcomes must be willing to route finance through voluntary channels rather than demanding state coercion, and the broader public must contribute voluntarily in sufficient volumes to reach conservation targets. Both conditions are empirically open questions.

We test them using a most-likely case design. Sweden ranks among the world's highest for social trust [@WVS2022], environmental identity, and per-capita conservation spending [@OECD2020] — the most structurally favorable context for voluntary instrument support globally. An important distinction sharpens the design logic: individual environmental *willingness* (the inclination to act voluntarily) and *structural instrument preference* (whether one prefers voluntary or mandatory policy architecture) are conceptually distinct. Sweden is the ceiling for the former. We test the latter. Even when individual willingness is maximized, do citizens prefer to route conservation finance through voluntary or mandatory channels?

We combine a preregistered discrete choice experiment (DCE, $n = 2{,}101$) with an incentivised donation experiment. The DCE presents paired choices between biodiversity programmes differing in financing instrument (tax, donation, certification, mandatory offsetting), land ownership type, monitoring regime, and annual cost. The donation experiment provides a revealed-preference measure of personal giving to a conservation NGO. We construct a respondent-level voluntary preference (VP) score from DCE choices and ask whether the most environmentally motivated and personally generous citizens endorse voluntary instruments or demand mandatory ones.

# Results

## Citizens prefer mandatory biodiversity financing

The conditional logit reveals a clear preference ordering (Fig. 1A): corporate offsetting carries the highest mean utility relative to tax (baseline; $\hat\beta_{\text{off}} = +0.146$, $p < 0.001$), followed by tax, market certification ($\hat\beta_{\text{cert}} = -0.082$, $p < 0.05$), and voluntary donation ($\hat\beta_{\text{don}} = -0.376$, $p < 0.001$). The opt-out (status quo, no programme) is strongly rejected (ASC $= -3.28$, $p < 0.001$). Results replicate in the mixed logit (Extended Data Table 1), which additionally reveals substantial preference heterogeneity: 31.3% of respondents prefer donation-financed programmes over tax and 45.2% prefer certification — but the mean preference is robustly mandatory. Public monitoring commands a willingness-to-pay (WTP) premium of approximately SEK 1,600/year; donation-financed programmes carry a WTP penalty of approximately SEK 1,930/year relative to the tax baseline (Fig. 1B).

The preference for corporate offsetting over tax reveals a within-mandatory distinction: beyond preferring mandatory instruments in general, Swedish citizens specifically favour locating the compliance burden with corporate actors. This is consistent with a polluter-pays norm and with the trust gradient in our data: respondents with higher trust in private markets show a larger offsetting WTP premium (SEK 1,117/year, high market trust; SEK 712/year, low market trust), while preference for tax over voluntary instruments is stable across all trust groups.

## No strategic free-riding: effectiveness drives mandatory preference

To test whether apparent mandatory preference is driven by strategic free-riding (high-concern respondents endorsing mandatory instruments so others are forced to pay while they benefit), we construct a respondent-level VP score — the share of mixed-instrument DCE tasks in which the respondent chose the voluntary option (mean = 0.431, SD = 0.320; Fig. 2A) — and regress it on donation behaviour, environmental concern, and their interaction.

Fractional logit (Papke-Wooldridge) with HC3 robust standard errors yields: APE$_{\text{don}} = -0.030$ ($p < 0.001$), APE$_{\text{env}} = -0.042$ ($p < 0.001$), interaction APE $= -0.002$ ($p = 0.792$). Both donation behaviour and environmental concern independently predict stronger mandatory preference; they do not interact (Fig. 2B).

The 2×2 cross-classification of environmental concern and donation behaviour (Fig. 3) rules out strategic free-riding directly. The strategic hypothesis predicts that the high-concern, no-donation cell produces the highest VP. The data show exactly the opposite: this cell has the lowest mean VP (0.385, 95% CI 0.317–0.453, $n = 84$), substantially below the low-concern, no-donation cell (0.549, 95% CI 0.502–0.596), with non-overlapping confidence intervals. The low-concern, no-donation cell reflects disengagement: respondents who neither care about biodiversity nor give are weakly indifferent between instruments. The high-concern, high-donation cell (VP = 0.392) reflects consequentialism: actual contributors who nonetheless demand mandatory schemes, because they recognise that voluntary participation cannot guarantee adequate provision at scale (assurance motive) or distrust private monitoring to deliver conservation outcomes (effectiveness concern). Both motivations resolve to the same instrument choice and are consistent with qualitative open-ended responses: *"Relying on donations means choosing not to care whether we reach the goals"*; *"Market mechanisms don't work; simply put, not enough money will be collected."*

Critically, the findings imply that voluntary instruments lack a natural constituency. The citizens who care most about biodiversity prefer mandatory instruments, and those who care least will not contribute voluntarily in sufficient volumes regardless of instrument design.

## Heterogeneity

Pre-registered subgroup models stratified by gender, income, and environmental attitude confirm the preference ordering is stable across groups (Extended Data Fig. 1). Exploratory analyses reveal that the political divide runs between state-coercive and market/corporate accountability mechanisms: left-bloc voters strongly prefer state tax, centre-right voters prefer corporate offsetting. Individual voluntary donation is rejected robustly by all political subgroups — the cross-partisan consensus is on mandatory mechanisms, not on which mandatory instrument. Low state trust shifts preferences within mandatory instruments (toward corporate offsetting, away from state tax) but does not produce voluntary instrument preference.

# Discussion

Swedish citizens consistently prefer mandatory biodiversity financing over voluntary alternatives, even in the context most structurally favorable to voluntary instrument support globally. The preference is not a statistical artefact: it replicates across conditional logit, mixed logit, latent class (Extended Data Table 2), and WTP specifications, and is corroborated by both qualitative responses and revealed donation behaviour. Crucially, the most environmentally committed and personally generous respondents most strongly prefer mandatory instruments, ruling out strategic free-riding. Voluntary instruments have no natural constituency: those who care most reject them on effectiveness grounds, and those who care least will not use them.

The finding carries a direct implication for GBF implementation. Target 19's reliance on voluntary private finance is misaligned with citizen preferences in the context most favorable to voluntary support. If mandatory preference persists in lower-trust contexts — as our theorized boundary condition implies — the voluntary governance turn lacks the citizen demand it implicitly assumes.

We theorize that voluntary instrument preference, where it appears, reflects a constrained defensive choice born of institutional weakness rather than unconstrained endorsement of voluntary governance. In high-trust, high-capacity state contexts, citizens display public-goods consequentialism: they recognise that voluntary participation cannot reach scale and prefer binding collective mechanisms. In lower-trust, lower-capacity contexts, environmentally concerned citizens may pragmatically prefer voluntary or NGO-led instruments as a bypass of a distrusted state — not because voluntary governance is structurally superior, but because mandatory mechanisms are perceived as unavailable or corrupted. Within-sample evidence is consistent: even Swedish respondents with the lowest state trust rank mandatory corporate offsetting first, not voluntary donation — suggesting the theorized inversion, if it occurs cross-nationally, is specific to contexts where both state capacity and private corporate accountability are absent.

This framing reinterprets the supply-side shift toward voluntary governance. The proliferation of voluntary biodiversity finance mechanisms under the GBF is frequently justified as matching decentralised citizen preferences. Our findings challenge this narrative at its most favorable test case. Voluntary instruments appear, in this light, as a second-best response to institutional failure — and devolving governance to voluntary architectures may be better understood as institutional resignation than as progressive policy design.

Future research should replicate these findings in lower-trust contexts to test the predicted preference inversion, ideally with a preregistered cross-national design. If mandatory preference persists across institutional contexts, the case for re-centering biodiversity finance around binding public mechanisms becomes considerably stronger.

# Methods

Full methods are provided in the Supplementary Information. Key elements are summarised here.

**Study design and preregistration.** Data were collected via online survey of Swedish citizens (Lysio Research, April–May 2026), preregistered prior to data collection (AsPredicted [blinded for review]). The final analysis sample comprises $n = 2{,}101$ respondents (48.8% female, median age 43), recruited to represent the Swedish adult population by age, gender, and region.

**Discrete choice experiment.** A $4 \times 3 \times 2 \times 4$ factorial design: financing instrument (tax, donation, certification, mandatory offsetting), land ownership (public, industrial private, small-scale private), monitoring (public administration, private audit), and annual cost (SEK 492–7,380/year). Each respondent completed 8 choice tasks (two programme options plus status quo). The VP score is the share of mixed-instrument tasks in which the respondent chose the voluntary option.

**Donation experiment.** Following the DCE, respondents allocated a portion of a potential SEK 1,000 prize to Naturskyddsföreningen. With 100 winners from 2,101 respondents, the game is incentive-compatible in expectation. Giving rate: 83.3%; median donation SEK 400.

**Statistical models.** Conditional logit and mixed logit (500 Halton draws, normal distributions on 6 random parameters) via logitr 1.2.0. WTP via delta method from the MXL Hessian. VP analysis: fractional logit (Papke-Wooldridge) with HC3 robust standard errors; average partial effects reported. Latent class MNL via Apollo 0.3.8. Confirmatory analyses preregistered; VP and subgroup analyses exploratory.

**Data and code availability.** Anonymized microdata and full replication code (R, Quarto) will be deposited upon acceptance. Open-ended responses (Q17) excluded from public deposit; coded subset available from the corresponding author. Ethics: Etikprövningsmyndigheten, ID 2025-04420-01.

# Figures

```{r}
#| label: fig-preferences
#| fig-cap: "Biodiversity financing instrument preferences. (A) Conditional logit coefficients relative to tax baseline (95% CI); n = 16,808 choice observations, 2,101 respondents. (B) Willingness to pay (WTP, SEK/year) from mixed logit delta-method estimates (95% CI). Mandatory instruments (offsetting, tax) shown in blue; voluntary instruments (certification, donation) in orange."
#| fig-width: 6.5
#| fig-height: 5.5
#| eval: true

mxl <- readRDS(here::here("analysis", "MXL_full_logitr.rds"))
cf   <- coef(mxl)
ses  <- logitr::se(mxl)
V    <- solve(-mxl$hessian)
p_c  <- cf["price"]

wtp_delta <- function(attr_nm) {
  b  <- cf[attr_nm]
  ai <- which(names(cf) == attr_nm)
  pi <- which(names(cf) == "price")
  g  <- numeric(length(cf))
  g[ai] <-  1 / (-p_c) * 1000
  g[pi] <- -b / (p_c^2) * 1000
  list(wtp = b / (-p_c) * 1000,
       se  = sqrt(abs(as.numeric(t(g) %*% V %*% g))))
}

wtp_df <- tibble(
  attr  = c("donation_mean", "cert_mean", "off_mean",
            "ind_mean", "pland_mean", "pmon_mean"),
  label = c("Donation", "Certification", "Offsetting",
            "Industrial land", "Public land", "Public monitoring"),
  type  = c("voluntary", "voluntary", "mandatory",
            "land", "land", "monitoring")
) |>
  rowwise() |>
  mutate(wtp = wtp_delta(attr)$wtp,
         se  = wtp_delta(attr)$se,
         lo  = wtp - 1.96 * se,
         hi  = wtp + 1.96 * se) |>
  ungroup()

instr_colors <- c("mandatory" = "#0072B2", "voluntary" = "#D55E00",
                  "land" = "#888888", "monitoring" = "#888888")

cl_coefs <- tibble(
  label = c("Donation", "Certification", "Offsetting",
            "Industrial land", "Public land", "Public monitoring"),
  coef  = c(-0.376, -0.082, 0.146, 0.010, 0.190, 0.315),
  se    = c(0.032, 0.032, 0.034, 0.030, 0.028, 0.024),
  type  = c("voluntary", "voluntary", "mandatory", "land", "land", "monitoring")
) |>
  mutate(lo = coef - 1.96 * se, hi = coef + 1.96 * se,
         label = factor(label, levels = rev(c("Donation", "Certification",
           "Offsetting", "Industrial land", "Public land", "Public monitoring"))))

p_coef <- ggplot(cl_coefs, aes(x = coef, y = label, colour = type)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_linerange(aes(xmin = lo, xmax = hi), linewidth = 0.8) +
  geom_point(size = 3) +
  scale_colour_manual(values = instr_colors, guide = "none") +
  labs(x = "Coefficient (vs. tax baseline)", y = NULL,
       title = "A  Instrument preferences (CL)") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", size = 10))

wtp_plot_df <- wtp_df |>
  filter(type %in% c("mandatory", "voluntary")) |>
  mutate(label = factor(label, levels = rev(c("Donation", "Certification", "Offsetting"))))

p_wtp <- ggplot(wtp_plot_df, aes(x = wtp, y = label, colour = type)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_linerange(aes(xmin = lo, xmax = hi), linewidth = 0.8) +
  geom_point(size = 3) +
  scale_colour_manual(values = instr_colors, guide = "none") +
  labs(x = "WTP relative to tax (SEK/year)", y = NULL,
       title = "B  Willingness to pay (MXL)") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", size = 10))

p_coef / p_wtp
```

```{r}
#| label: fig-vp
#| fig-cap: "Voluntary preference (VP) score analysis. (A) Distribution of VP scores (share of mixed-instrument tasks where voluntary option was chosen). Blue: lean mandatory (VP ≤ 0.5, 67%); orange: lean voluntary (VP > 0.5, 33%). (B) VP score by donation amount, stratified by environmental attitude tercile (OLS fits with 95% CI). Convergent slopes indicate no interaction; the null interaction (p = 0.792) rules out strategic free-riding."
#| fig-width: 6.5
#| fig-height: 5.5
#| eval: true

vp_mean  <- mean(anal_data$voluntary_pref)
pct_mand <- round(mean(anal_data$voluntary_pref <= 0.5) * 100)
pct_vol  <- round(mean(anal_data$voluntary_pref >  0.5) * 100)

dens     <- density(anal_data$voluntary_pref, from = 0, to = 1, n = 512)
dens_df  <- tibble(x = dens$x, y = dens$y)
y_at_half <- approx(dens$x, dens$y, xout = 0.5)$y
y_label  <- max(dens$y) * 0.45
dens_mand <- dens_df |> filter(x <= 0.5) |> add_row(x = 0.5, y = y_at_half) |> arrange(x)
dens_vol  <- dens_df |> filter(x >= 0.5) |> add_row(x = 0.5, y = y_at_half) |> arrange(x)

p_dist <- ggplot() +
  geom_area(data = dens_mand, aes(x = x, y = y), fill = "#0072B2", alpha = 0.55) +
  geom_area(data = dens_vol,  aes(x = x, y = y), fill = "#D55E00", alpha = 0.55) +
  geom_line(data = dens_df,   aes(x = x, y = y), colour = "grey20", linewidth = 0.5) +
  geom_vline(xintercept = 0.5, colour = "grey30", linewidth = 0.7) +
  geom_vline(xintercept = vp_mean, linetype = "dashed", colour = "grey30", linewidth = 0.55) +
  annotate("text", x = 0.20, y = y_label,
           label = paste0(pct_mand, "% lean\nmandatory"),
           colour = "#0072B2", size = 3.2, fontface = "bold", hjust = 0.5) +
  annotate("text", x = 0.78, y = y_label,
           label = paste0(pct_vol, "% lean\nvoluntary"),
           colour = "#D55E00", size = 3.2, fontface = "bold", hjust = 0.5) +
  scale_x_continuous(breaks = c(0, 0.25, 0.5, 0.75, 1),
    labels = c("0\n(always\nmandatory)", "0.25", "0.5", "0.75", "1\n(always\nvoluntary)"),
    limits = c(-0.01, 1.01)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.06))) +
  labs(x = "Voluntary preference score (VP)", y = "Density",
       title = "A  VP score distribution") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank(), panel.grid.major.x = element_blank(),
        axis.text.x = element_text(size = 8, lineheight = 0.85),
        plot.title = element_text(face = "bold", size = 10))

p_interact <- ggplot(anal_data, aes(x = donation, y = voluntary_pref, colour = env_lab)) +
  geom_jitter(alpha = 0.12, size = 0.7, width = 25, height = 0.02) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1) +
  scale_colour_manual(
    values = c("Low env. att." = "#D55E00", "Medium" = "#E69F00",
               "High env. att." = "#0072B2"),
    name = "Environmental\nattitude") +
  scale_x_continuous(labels = scales::label_comma()) +
  scale_y_continuous(labels = scales::percent, limits = c(-0.1, 1.1)) +
  labs(x = "Donation (SEK)", y = "Voluntary preference (VP)",
       title = "B  VP by donation × environmental attitude") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold", size = 10))

p_dist / p_interact
```

```{r}
#| label: fig-2x2
#| fig-cap: "Mean voluntary preference (VP) score by environmental attitude (median split on Q11) and donation behaviour (donated > 0 vs. zero). Error bars: 95% CI. The strategic free-rider hypothesis predicts highest VP in the high-concern, no-donation cell; the data show exactly the opposite. n in each cell shown."
#| fig-width: 5
#| fig-height: 3.8
#| eval: true

twox2 <- anal_data |>
  mutate(
    env_grp = if_else(env_hi, "High env. concern", "Low env. concern"),
    don_grp = if_else(donated, "Donated", "Did not donate")
  ) |>
  group_by(env_grp, don_grp) |>
  summarise(
    mean_vp = mean(voluntary_pref),
    se_vp   = sd(voluntary_pref) / sqrt(n()),
    n       = n(),
    .groups = "drop"
  ) |>
  mutate(
    ci_lo   = mean_vp - 1.96 * se_vp,
    ci_hi   = mean_vp + 1.96 * se_vp,
    env_grp = factor(env_grp, levels = c("Low env. concern", "High env. concern")),
    don_grp = factor(don_grp, levels = c("Did not donate", "Donated")),
    fill    = case_when(
      env_grp == "High env. concern" & don_grp == "Did not donate" ~ "focal",
      TRUE ~ "other"
    )
  )

ggplot(twox2, aes(x = don_grp, y = mean_vp, fill = fill)) +
  geom_col(width = 0.55, alpha = 0.85) +
  geom_errorbar(aes(ymin = ci_lo, ymax = ci_hi), width = 0.15, colour = "grey30") +
  geom_text(aes(label = paste0("n = ", n), y = ci_hi + 0.02),
            size = 2.8, colour = "grey40") +
  facet_wrap(~env_grp) +
  scale_fill_manual(values = c("focal" = "#0072B2", "other" = "#888888"),
                    guide = "none") +
  scale_y_continuous(labels = scales::percent, limits = c(0, 0.70)) +
  labs(x = NULL, y = "Mean VP score",
       caption = "Blue bar = focal strategic free-rider cell (lowest VP, not highest)") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank(),
        strip.text = element_text(face = "bold"))
```

```{r}
#| label: fig-subgroups
#| fig-cap: "Instrument preference heterogeneity by political bloc (exploratory, not preregistered). Coefficients from subgroup conditional logit models relative to tax baseline (95% CI). Individual voluntary donation is rejected by all political blocs; the cross-partisan divide is between state tax and corporate offsetting."
#| fig-width: 6.5
#| fig-height: 4.5
#| eval: true

all_exp <- readRDS(here::here("analysis", "heterogeneity_exploratory.rds"))

instr_colors2 <- c(
  "Donation"      = "#D55E00",
  "Certification" = "#E69F00",
  "Offsetting"    = "#0072B2"
)

party_order <- c("Left (S/V/MP)", "Centre-right", "SD", "No pref / DK")

all_exp |>
  filter(Label %in% party_order) |>
  mutate(
    ci_lo      = Coef - 1.96 * SE,
    ci_hi      = Coef + 1.96 * SE,
    Instrument = recode(Attr,
      don  = "Donation",
      cert = "Certification",
      off  = "Offsetting"
    ),
    Label = factor(Label, levels = party_order)
  ) |>
  filter(Instrument %in% c("Donation", "Certification", "Offsetting")) |>
  ggplot(aes(x = Coef, y = Label, colour = Instrument)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_linerange(aes(xmin = ci_lo, xmax = ci_hi),
                 position = position_dodge(width = 0.5), linewidth = 0.8) +
  geom_point(position = position_dodge(width = 0.5), size = 3) +
  scale_colour_manual(values = instr_colors2, name = "Instrument") +
  labs(x = "Coefficient vs. tax baseline", y = NULL,
       title = "Instrument preferences by political bloc") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank(),
        legend.position = "bottom",
        plot.title = element_text(face = "bold", size = 10))
```

# References

::: {#refs}
:::
