---
title: "Supplementary Information"
subtitle: "Citizens prefer mandatory over voluntary biodiversity financing even under favorable conditions"
format:
  pdf:
    number-sections: false
    fontsize: 11pt
    geometry: margin=2.5cm
    keep-tex: false
    include-in-header:
      text: |
        % Supplementary numbering: Table S1..., Figure S1...
        \renewcommand{\thetable}{S\arabic{table}}
        \renewcommand{\thefigure}{S\arabic{figure}}
        % Flush caption body instead of KOMA's hanging indent
        \setcapindent{0pt}
bibliography: references.bib
csl: nature.csl
execute:
  eval: true
  echo: false
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(knitr)
library(kableExtra)
library(here)
library(readxl)
library(sandwich)
library(lmtest)
library(logitr)
```

```{r load-data}
# Load anonymised microdata from public deposit (pid and Q17 excluded)
full_data <- read_csv(here("deposit", "biodivfinance_microdata_anon.csv"),
                      show_col_types = FALSE)

# Load DCE design file (public, in repository)
price_map  <- c("0" = 492, "1" = 2460, "2" = 4920, "3" = 7380)
design_raw <- read_excel(here("design", "Trial 3 - factorial grouped, svensk.xlsx"), sheet = 1)
names(design_raw) <- c(
  "cs", "block", "a_source", "a_src_txt", "a_land", "a_land_txt",
  "a_monitor", "a_mon_txt", "a_price", "a_price_txt",
  "b_source", "b_src_txt", "b_land", "b_land_txt",
  "b_monitor", "b_mon_txt", "b_price", "b_price_txt"
)
design <- design_raw |>
  select(cs, block, a_source, a_land, a_monitor, a_price,
         b_source, b_land, b_monitor, b_price) |>
  group_by(block) |> mutate(task = row_number()) |> ungroup() |>
  mutate(a_price_sek = price_map[as.character(a_price)],
         b_price_sek = price_map[as.character(b_price)])

# Reconstruct choice data in long format (mirrors mxl_analysis.R pipeline)
choices_ord1 <- full_data |> filter(ordning == 1) |>
  select(id, block, choice1:choice8) |>
  pivot_longer(choice1:choice8, names_to = "col", values_to = "choice") |>
  mutate(task = as.integer(substr(col, 7, 7)))
choices_ord2 <- full_data |> filter(ordning == 2) |>
  select(id, block, choice12, choice22, choice32, choice42,
         choice52, choice62, choice72, choice82) |>
  pivot_longer(-c(id, block), names_to = "col", values_to = "choice") |>
  mutate(task = as.integer(substr(col, 7, 7)))

resp <- bind_rows(choices_ord1, choices_ord2) |>
  filter(!is.na(choice)) |>
  left_join(design, by = c("block", "task")) |>
  mutate(choice_num = case_when(choice == "a" ~ 1L, choice == "b" ~ 2L, TRUE ~ 3L))
```

```{r build-long}
# Build 3-row-per-task long format for CL estimation
obs_int <- as.integer(factor(paste0(resp$id, "_", resp$task)))

df_a <- data.frame(
  obsID  = obs_int, respID = resp$id, altID = "a",
  chosen = as.integer(resp$choice_num == 1),
  don    = as.integer(resp$a_source == 1),
  cert   = as.integer(resp$a_source == 2),
  off    = as.integer(resp$a_source == 3),
  ind    = as.integer(resp$a_land == 1),
  pland  = as.integer(resp$a_land == 2),
  pmon   = as.integer(resp$a_monitor == 1),
  price  = resp$a_price_sek / 1000   # price in thousands SEK
)
df_b <- data.frame(
  obsID  = obs_int, respID = resp$id, altID = "b",
  chosen = as.integer(resp$choice_num == 2),
  don    = as.integer(resp$b_source == 1),
  cert   = as.integer(resp$b_source == 2),
  off    = as.integer(resp$b_source == 3),
  ind    = as.integer(resp$b_land == 1),
  pland  = as.integer(resp$b_land == 2),
  pmon   = as.integer(resp$b_monitor == 1),
  price  = resp$b_price_sek / 1000
)
df_n <- data.frame(
  obsID  = obs_int, respID = resp$id, altID = "neither",
  chosen = as.integer(resp$choice_num == 3),
  don = 0, cert = 0, off = 0, ind = 0, pland = 0, pmon = 0, price = 0
)
long_db <- rbind(df_a, df_b, df_n)
long_db <- long_db[order(long_db$respID, long_db$obsID, long_db$altID), ]
long_db$asc_neither <- as.integer(long_db$altID == "neither")
```

```{r vp-data}
# Construct VP scores (same logic as instrument_pref_analysis.R)
resp_scored <- resp |>
  mutate(
    a_vol      = a_source %in% c(1L, 2L),   # donation or certification
    b_vol      = b_source %in% c(1L, 2L),
    a_man      = a_source %in% c(0L, 3L),   # tax or offsetting
    b_man      = b_source %in% c(0L, 3L),
    mixed_task = (a_vol & b_man) | (a_man & b_vol),
    chose_vol  = case_when(
      !mixed_task | choice == "3" ~ NA_real_,
      (a_vol & choice == "a") | (b_vol & choice == "b") ~ 1,
      TRUE ~ 0
    )
  )

resp_level <- resp_scored |>
  filter(mixed_task, !is.na(chose_vol)) |>
  group_by(id) |>
  summarise(voluntary_pref = mean(chose_vol), n_mixed = n(), .groups = "drop")

covars <- full_data |>
  transmute(
    id,
    donation    = yourdonation,
    gave        = as.integer(yourdonation > 0),
    env_import  = q11,
    bio_knowl   = q12,
    trust_ngo   = q2_3,
    trust_state = q2_1,
    policy_cq   = q14,
    income      = q8,
    gender      = q10
  )

anal_data <- resp_level |>
  left_join(covars, by = "id") |>
  filter(!is.na(donation), !is.na(env_import), !is.na(income), !is.na(gender)) |>
  mutate(
    donation_z  = scale(donation)[, 1],
    env_z       = scale(env_import)[, 1],
    female      = as.integer(gender == 2),
    income_mid  = as.integer(income == 2),
    income_high = as.integer(income >= 3),
    gave_binary = gave
  )
```

# Supplementary Methods

## S1. Discrete choice experiment design

Each respondent completed eight choice tasks. Every task presented two hypothetical biodiversity programme alternatives plus a status-quo option ("neither programme"). Programmes were described by four attributes whose levels are given in Table S1. The $4 \times 3 \times 2 \times 4$ full factorial was reduced to a blocked fractional factorial design (one block per respondent, 8 tasks per block). Within each task, attribute levels for options A and B were drawn independently from the design.

```{r attr-table}
attr_tab <- tribble(
  ~Attribute,            ~Level,                                       ~Note,
  "Financing instrument", "Tax (public budget allocation)",            "Baseline",
  "",                     "Voluntary private donation",                "",
  "",                     "Consumer certification scheme",             "",
  "",                     "Mandatory corporate offsetting",            "",
  "Land ownership",       "Small-scale private land",                  "Baseline",
  "",                     "Industrial private land",                   "",
  "",                     "Public land",                               "",
  "Monitoring",           "Private audit",                             "Baseline",
  "",                     "Public administration",                     "",
  "Annual household cost","SEK 492/year (approx. USD 49)",             "",
  "",                     "SEK 2,460/year (approx. USD 246)",          "",
  "",                     "SEK 4,920/year (approx. USD 492)",          "",
  "",                     "SEK 7,380/year (approx. USD 738)",          ""
)

kbl(attr_tab, booktabs = TRUE, longtable = FALSE, linesep = "",
    caption = "DCE attribute levels. Baseline levels serve as the reference category in conditional and mixed logit models.",
    col.names = c("Attribute", "Level", "Note"),
    align = "llc") |>
  kable_styling(full_width = TRUE, latex_options = "hold_position") |>
  column_spec(1, bold = TRUE)
```

**Hypothetical-bias and cost-vector mitigation.** Before the choice tasks,
respondents were shown a budget-substitution reminder: "In your answers to the
subsequent choices, please consider that the money that you spend on
biodiversity conservation will no longer be available for other purposes." The
opt-out was explained explicitly ("If you choose 'none of the above', it implies
that everything stays as it is (status quo), and no financing option is
implemented"), and respondents were instructed to "answer as if you actually had
to incur the annual cost for your household associated with your choice."
Combined cheap-talk and opt-out reminder devices of this kind have been shown to
reduce cost-vector effects on welfare estimates substantially [@Borger2025].

**Instrument coding.** Financing source was coded relative to the tax baseline: donation ($\hat\beta_\text{don}$), certification ($\hat\beta_\text{cert}$), and mandatory offsetting ($\hat\beta_\text{off}$). Land ownership was coded relative to small-scale private land: industrial private ($\hat\beta_\text{ind}$) and public ($\hat\beta_\text{pland}$). Monitoring was coded relative to private audit: public administration ($\hat\beta_\text{pmon}$). Price entered the utility function in thousands SEK to improve numerical scaling.

**Alternative-specific constant.** An alternative-specific constant ($\text{ASC}_\text{neither}$) was included for the opt-out (status-quo) alternative. A negative ASC indicates a systematic preference for any programme over the status quo.

## S2. Voluntary preference (VP) score

The VP score quantifies each respondent's tendency to choose voluntary over mandatory financing instruments in the DCE. Only tasks in which the two programme alternatives presented one voluntary instrument (donation or certification) and one mandatory instrument (tax or mandatory offsetting) are included; tasks where both options share the same instrument type, or where the respondent chose the opt-out, are excluded from the denominator. For eligible "mixed-instrument" tasks, VP equals the proportion in which the respondent chose the voluntary option, ranging from 0 (always mandatory) to 1 (always voluntary). The mean number of mixed-instrument tasks per respondent was `r round(mean(anal_data$n_mixed), 1)` (range `r min(anal_data$n_mixed)`–`r max(anal_data$n_mixed)`).

## S3. Donation experiment

Immediately following the DCE, respondents allocated a share of a potential SEK 1,000 prize to Naturskyddsföreningen (the Swedish Society for Nature Conservation). One hundred winners were drawn randomly from all $n = 2{,}101$ respondents, making the game incentive-compatible in expectation: each respondent faced an approximately 1-in-21 chance of their allocation being paid out. The donation amount (SEK 0–1,000) is used as a continuous measure of individual propensity to contribute to biodiversity conservation voluntarily and as a binary indicator (gave > 0).

## S4. Sample characterisation

```{r sample-table}
n_total <- nrow(full_data)

gender_n <- full_data |>
  mutate(g = case_when(q10==1~"Male", q10==2~"Female", TRUE~"Other/NR")) |>
  count(g)

party_pct <- full_data |>
  mutate(party = case_when(
    q16 == 1  ~ "Social Democrats (S)",
    q16 == 2  ~ "Sweden Democrats (SD)",
    q16 == 3  ~ "Moderate Party (M)",
    q16 == 4  ~ "Green Party (MP)",
    q16 == 5  ~ "Centre Party (C)",
    q16 == 6  ~ "Christian Democrats (KD)",
    q16 == 7  ~ "Left Party (V)",
    q16 == 8  ~ "Liberal Party (L)",
    q16 == 9  ~ "Other party",
    q16 == 10 ~ "No preference / DK"
  )) |>
  count(party) |>
  arrange(desc(n)) |>
  mutate(label = paste0(party, ": ", n, " (", round(n/n_total*100,1), "%)"))

fem_n   <- filter(gender_n, g == "Female")$n
male_n  <- filter(gender_n, g == "Male")$n
other_n <- filter(gender_n, g == "Other/NR")$n

# escape = FALSE below (needed for $\bar{x}$ math), so % must be escaped as \%
# manually — an unescaped % starts a LaTeX comment and swallows the row break
pct <- function(x) paste0(round(x, 1), "\\%")

desc_tab <- tribble(
  ~Characteristic, ~Value,
  "Analysis sample ($n$)",          format(n_total, big.mark = ","),
  "Female",                          paste0(fem_n,   " (", pct(fem_n/n_total*100),   ")"),
  "Male",                            paste0(male_n,  " (", pct(male_n/n_total*100),  ")"),
  "Other / prefer not to say",       paste0(other_n, " (", pct(other_n/n_total*100), ")"),
  "Env. importance (Q11; 1--5 scale)",
    paste0("Mean $\\bar{x}$ = ", round(mean(full_data$q11, na.rm=TRUE),2),
           ", SD = ", round(sd(full_data$q11, na.rm=TRUE),2)),
  "Biodiversity knowledge (Q12; 1--5 scale)",
    paste0("Mean $\\bar{x}$ = ", round(mean(full_data$q12, na.rm=TRUE),2),
           ", SD = ", round(sd(full_data$q12, na.rm=TRUE),2)),
  "Donation amount (SEK)",           paste0("Mean = ", round(mean(full_data$yourdonation),0),
                                           ", Median = ", median(full_data$yourdonation),
                                           ", SD = ", round(sd(full_data$yourdonation),0)),
  "Donated $>$ 0",                   paste0(sum(full_data$yourdonation>0),
                                           " (", pct(mean(full_data$yourdonation>0)*100), ")")
)

kbl(desc_tab, booktabs = TRUE, longtable = FALSE, escape = FALSE, linesep = "",
    caption = "Sample descriptive statistics.",
    col.names = c("Characteristic", "Value"),
    align = "ll") |>
  kable_styling(full_width = TRUE, latex_options = "hold_position")
```

Party composition: Social Democrats 23.6%, Sweden Democrats 13.7%, Moderate Party 14.6%, Green Party 7.8%, Centre Party 4.8%, Christian Democrats 3.7%, Left Party 6.5%, Liberal Party 1.9%, Other 7.2%, No preference or don't know 16.3%. Relative to the 2022 Riksdag election, Sweden Democrats (13.7% vs. 20.5% vote share) and Social Democrats (23.6% vs. 30.3%) are under-represented; smaller parties are broadly proportional. As SD voters show the weakest mandatory preference (main text Fig. 5), their under-representation modestly biases the sample's lean-mandatory VP share upward; within-sample subgroup findings are unaffected.

# Supplementary Results

## Subgroup heterogeneity

Pre-registered subgroup conditional logit models stratified by gender, income tercile, and environmental attitude tercile are shown in Fig. S1. The mandatory-over-voluntary ordering holds in every subgroup: donation is negative throughout, and mandatory offsetting is positive or near zero. The one notable gradient is by environmental attitude: the preference against voluntary donation strengthens monotonically from the lowest to the highest attitude tercile, consistent with the effectiveness-concern interpretation developed in the main text.

```{r}
#| label: fig-heterogeneity
#| fig-cap: "Instrument preference heterogeneity by pre-registered subgroups. Coefficients from subgroup conditional logit models relative to tax baseline (95% CI), stratified by gender, income tercile, and environmental attitude tercile. Only the three financing instrument coefficients (donation, certification, offsetting) are shown; land and monitoring coefficients are omitted for clarity. The mandatory-over-voluntary preference ordering is broadly consistent across all subgroups."
#| fig-width: 7
#| fig-height: 5
#| eval: true

# Subgroup CL helper — runs from deposit data loaded above
make_long_sub <- function(data) {
  obs <- as.integer(factor(paste0(data$id, "_", data$task)))
  bind_rows(
    data.frame(obsID=obs, chosen=as.integer(data$choice_num==1), asc=0L,
               don=as.integer(data$a_source==1), cert=as.integer(data$a_source==2),
               off=as.integer(data$a_source==3), ind=as.integer(data$a_land==1),
               pland=as.integer(data$a_land==2), pmon=as.integer(data$a_monitor==1),
               price=data$a_price_sek/1000),
    data.frame(obsID=obs, chosen=as.integer(data$choice_num==2), asc=0L,
               don=as.integer(data$b_source==1), cert=as.integer(data$b_source==2),
               off=as.integer(data$b_source==3), ind=as.integer(data$b_land==1),
               pland=as.integer(data$b_land==2), pmon=as.integer(data$b_monitor==1),
               price=data$b_price_sek/1000),
    data.frame(obsID=obs, chosen=as.integer(data$choice_num==3), asc=1L,
               don=0L, cert=0L, off=0L, ind=0L, pland=0L, pmon=0L, price=0)
  ) |> arrange(obsID)
}

run_cl_sub <- function(data, label) {
  long <- make_long_sub(data)
  m <- logitr(data=long, outcome="chosen", obsID="obsID",
               pars=c("asc","don","cert","off","ind","pland","pmon","price"),
               numMultiStarts=3)
  cf <- coef(m); se <- logitr::se(m)
  tibble(group=label,
         attr=c("don","cert","off"),
         coef=cf[c("don","cert","off")],
         se=se[c("don","cert","off")]) |>
    mutate(lo=coef-1.96*se, hi=coef+1.96*se)
}

# Join covariates needed for subgroup splits.
#
# Terciles MUST be assigned at the respondent level, not on the choice-task
# frame: `resp` holds 8 rows per respondent, so calling ntile() on it ranks
# 16,808 task rows rather than 2,101 people. Because ntile() breaks ties by row
# position, respondents sitting exactly on a tercile boundary would then have
# their 8 tasks split across two terciles and be counted in both subgroups.
resp_tercile <- full_data |>
  transmute(
    id,
    gender_grp = case_when(q10 == 1 ~ "Male", q10 == 2 ~ "Female",
                           TRUE ~ NA_character_),
    income_t   = ntile(q8,  3),
    env_t      = ntile(q11, 3)
  )

resp_cov <- resp |>
  left_join(resp_tercile, by = "id") |>
  filter(!is.na(gender_grp))

sub_res <- bind_rows(
  run_cl_sub(filter(resp_cov, gender_grp=="Male"),   "Male"),
  run_cl_sub(filter(resp_cov, gender_grp=="Female"), "Female"),
  run_cl_sub(filter(resp_cov, income_t==1),          "Income: low"),
  run_cl_sub(filter(resp_cov, income_t==2),          "Income: mid"),
  run_cl_sub(filter(resp_cov, income_t==3),          "Income: high"),
  run_cl_sub(filter(resp_cov, env_t==1),             "Env. att.: low"),
  run_cl_sub(filter(resp_cov, env_t==2),             "Env. att.: mid"),
  run_cl_sub(filter(resp_cov, env_t==3),             "Env. att.: high")
)

group_order <- c("Male","Female",
                 "Income: low","Income: mid","Income: high",
                 "Env. att.: low","Env. att.: mid","Env. att.: high")

instr_cols <- c(Donation="#D55E00", Certification="#E69F00", Offsetting="#3b4994")

sub_res |>
  mutate(
    Instrument = recode(attr, don="Donation", cert="Certification", off="Offsetting"),
    group = factor(group, levels=rev(group_order))
  ) |>
  ggplot(aes(x=coef, y=group, colour=Instrument)) +
  geom_vline(xintercept=0, linetype="dashed", colour="grey60") +
  geom_linerange(aes(xmin=lo, xmax=hi),
                 position=position_dodge(width=0.55), linewidth=0.7) +
  geom_point(position=position_dodge(width=0.55), size=2.8) +
  scale_colour_manual(values=instr_cols, name="Instrument") +
  labs(x="Coefficient vs. tax baseline", y=NULL) +
  theme_minimal(base_size=11) +
  theme(panel.grid.minor=element_blank(), legend.position="bottom")
```

## Choice model estimates

Full output for the conditional logit [@McFadden1974] is given in Table S3 and for the mixed logit [@McFaddenTrain2000; @Train2009] in Table S4. Both were estimated with logitr 1.2.0 [@Helveston2023]. The mixed logit uses 500 Sobol draws, normal mixing distributions on the six non-price attributes, and 10 random starts with a fixed seed to guard against local optima; the reported model is the best-fitting start.

```{r cl-table}
cl <- logitr(
  data    = long_db,
  outcome = "chosen",
  obsID   = "obsID",
  pars    = c("asc_neither","don","cert","off","ind","pland","pmon","price"),
  numMultiStarts = 5
)

cf  <- coef(cl)
se  <- logitr::se(cl)
zv  <- cf / se
pv  <- 2 * pnorm(abs(zv), lower.tail = FALSE)

par_nms  <- c("asc_neither","don","cert","off","ind","pland","pmon","price")
par_labs <- c("ASC (status quo / opt-out)", "Donation", "Certification",
              "Offsetting", "Industrial land", "Public land",
              "Public monitoring", "Price (TSEK/year)")

cl_tab <- tibble(
  Parameter = par_labs,
  Estimate  = sprintf("%.4f", cf[par_nms]),
  SE        = sprintf("%.4f", se[par_nms]),
  z         = sprintf("%.3f", zv[par_nms]),
  p         = ifelse(pv[par_nms] < 0.001, "<0.001", sprintf("%.4f", pv[par_nms]))
)

kbl(cl_tab, booktabs = TRUE, longtable = FALSE, align = "lrrrr", linesep = "",
    caption = paste0(
      "Conditional logit full model output. ",
      "Baseline: tax financing, small-scale private land, private monitoring. ",
      "$N = ", format(n_distinct(long_db$respID), big.mark=","), "$ respondents; ",
      format(n_distinct(long_db$obsID), big.mark=","), " choice observations. ",
      "Log-likelihood $= ", round(as.numeric(logLik(cl)), 2), "$; ",
      "AIC $= ", round(AIC(cl), 1), "$."
    )) |>
  kable_styling(full_width = TRUE, latex_options = "hold_position") |>
  column_spec(1, width = "5.5cm") |>
  footnote(
    general = "Price coefficient is per thousand SEK/year. Conditional logit WTP = -(attribute coefficient) / (price coefficient / 1000).",
    threeparttable = TRUE
  )
```


```{r mxl-table}
# Load fitted MXL model from deposit (model object; no individual-level data stored here
# beyond what is reconstructed from the public deposit CSV)
mxl <- readRDS(here("deposit", "MXL_full_logitr.rds"))

cf_m  <- coef(mxl)
se_m  <- logitr::se(mxl)
zv_m  <- cf_m / se_m
pv_m  <- 2 * pnorm(abs(zv_m), lower.tail = FALSE)

pars <- c("asc_neither","don","cert","off","ind","pland","pmon","price",
          "sd_don","sd_cert","sd_off","sd_ind","sd_pland","sd_pmon")
labs <- c(
  "Mean: ASC (opt-out)",
  "Mean: Donation", "Mean: Certification", "Mean: Offsetting",
  "Mean: Industrial land", "Mean: Public land", "Mean: Public monitoring",
  "Mean: Price (TSEK/year)",
  "SD: Donation", "SD: Certification", "SD: Offsetting",
  "SD: Industrial land", "SD: Public land", "SD: Public monitoring"
)

mxl_tab <- tibble(
  Parameter = labs,
  Estimate  = sprintf("%.4f", cf_m[pars]),
  SE        = sprintf("%.4f", se_m[pars]),
  z         = sprintf("%.3f", zv_m[pars]),
  p         = ifelse(pv_m[pars] < 0.001, "<0.001", sprintf("%.4f", pv_m[pars]))
)

kbl(mxl_tab, booktabs = TRUE, longtable = FALSE, align = "lrrrr", linesep = "",
    caption = paste0(
      "Mixed logit full model output. ",
      "Normal mixing distributions on 6 random parameters (donation, certification, ",
      "offsetting, industrial land, public land, public monitoring); ",
      "500 Sobol draws; 10 random starts; seed = 42. ",
      "Baseline: tax financing, small-scale private land, private monitoring. ",
      "Log-likelihood $= ", round(as.numeric(logLik(mxl)), 2), "$."
    )) |>
  kable_styling(full_width = TRUE, latex_options = "hold_position") |>
  column_spec(1, width = "5.5cm") |>
  footnote(
    general = "SD parameters capture preference heterogeneity across respondents. WTP via delta method from the MXL Hessian; see main text for WTP estimates.",
    threeparttable = TRUE
  )
```

## Voluntary preference regression

Table S5 reports the fractional logit (Papke-Wooldridge) models of the VP score. FP3 is the primary specification reported in the main text. FP4 substitutes a binary donated indicator for the continuous donation z-score as a robustness check. The donation × environmental importance interaction is small and not significant in either parameterisation, which is the result that speaks against the strategic free-riding account.

```{r vp-table}
# Fractional logit (Papke-Wooldridge) models
fp1 <- glm(voluntary_pref ~ donation_z + env_z,
           data = anal_data, family = quasibinomial(link = "logit"))
fp2 <- glm(voluntary_pref ~ donation_z * env_z,
           data = anal_data, family = quasibinomial(link = "logit"))
fp3 <- glm(voluntary_pref ~ donation_z * env_z +
             trust_ngo + trust_state + policy_cq +
             bio_knowl + female + income_mid + income_high,
           data = anal_data, family = quasibinomial(link = "logit"))
fp4 <- glm(voluntary_pref ~ gave_binary * env_z +
             trust_ngo + trust_state + policy_cq +
             bio_knowl + female + income_mid + income_high,
           data = anal_data, family = quasibinomial(link = "logit"))

# Average partial effect scale factor for FP3 (primary model)
ape_scale <- mean(fitted(fp3) * (1 - fitted(fp3)))

# Helper: formatted "est*** (se)" string
fmt_cell <- function(m, var) {
  ct <- coeftest(m, vcov = vcovHC(m, "HC3"))
  if (!var %in% rownames(ct)) return("")
  r   <- ct[var, ]
  est <- sprintf("%.3f", r["Estimate"])
  se  <- sprintf("%.3f", r["Std. Error"])
  p   <- r["Pr(>|z|)"]
  stars <- if (p < .001) "***" else if (p < .01) "**" else if (p < .05) "*" else if (p < .1) "." else ""
  paste0(est, stars, " (", se, ")")
}

dev_r2 <- function(m) sprintf("%.4f", 1 - m$deviance / m$null.deviance)

vars <- c("(Intercept)", "donation_z", "env_z", "donation_z:env_z",
          "gave_binary", "gave_binary:env_z",
          "trust_ngo", "trust_state", "policy_cq",
          "bio_knowl", "female", "income_mid", "income_high")
vlabs <- c("Intercept",
           "Donation (z-score)", "Env. importance (z-score)",
           "Donation $\\times$ Env. importance",
           "Donated (binary)", "Donated $\\times$ Env. importance",
           "Trust: NGOs (Q2\\_3)", "Trust: state (Q2\\_1)",
           "Policy consequentiality (Q14)",
           "Biodiversity knowledge (Q12)",
           "Female", "Income: mid", "Income: high")

tab_rows <- map2_dfr(vars, vlabs, function(v, l) {
  tibble(
    Variable = l,
    FP1 = fmt_cell(fp1, v),
    FP2 = fmt_cell(fp2, v),
    FP3 = fmt_cell(fp3, v),
    FP4 = fmt_cell(fp4, v)
  )
})

# APE for FP3 key predictors
ape_don <- sprintf("%.3f", coef(fp3)["donation_z"]   * ape_scale)
ape_env <- sprintf("%.3f", coef(fp3)["env_z"]        * ape_scale)
ape_int <- sprintf("%.3f", coef(fp3)["donation_z:env_z"] * ape_scale)

fit_rows <- tibble(
  Variable = c("Dev. $R^2$", "$N$"),
  FP1 = c(dev_r2(fp1), as.character(nobs(fp1))),
  FP2 = c(dev_r2(fp2), as.character(nobs(fp2))),
  FP3 = c(dev_r2(fp3), as.character(nobs(fp3))),
  FP4 = c(dev_r2(fp4), as.character(nobs(fp4)))
)

bind_rows(tab_rows, fit_rows) |>
  kbl(booktabs = TRUE, longtable = FALSE, escape = FALSE, align = "lcccc", linesep = "",
      caption = paste0(
        "Voluntary preference (VP) regression --- ",
        "fractional logit (Papke-Wooldridge). Dependent variable: VP score (0--1). ",
        "HC3 robust standard errors in parentheses. ",
        "FP1: main effects only (continuous donation z-score). ",
        "FP2: interaction added. ",
        "FP3: full model (primary specification, reported in main text). ",
        "FP4: binary donated indicator instead of continuous donation (robustness check). ",
        "Average partial effects for FP3: ",
        "donation z-score $= ", ape_don, "$; ",
        "env. importance z-score $= ", ape_env, "$; ",
        "interaction $= ", ape_int, "$."
      ),
      col.names = c("", "FP1", "FP2", "FP3", "FP4")) |>
  kable_styling(full_width = TRUE, latex_options = "hold_position",
                font_size = 9) |>
  column_spec(1, width = "4.5cm") |>
  footnote(
    general = "Significance: . $p<0.1$; * $p<0.05$; ** $p<0.01$; *** $p<0.001$.",
    escape  = FALSE,
    threeparttable = TRUE
  )
```

## Supplementary Table S6 | Perceived instrument effectiveness (Q13)

Question 13 asked respondents how effective each instrument is at delivering
biodiversity benefits, on a five-point scale, for four instruments: voluntary
donations (Q13\_1), market-based biodiversity certification (Q13\_2),
environmental taxes (Q13\_3) and mandatory offsetting (Q13\_4). Because Q13 was
administered *after* the choice tasks, it cannot identify the mechanism behind
observed choices: preference and perceived effectiveness may both reflect a
prior disposition, and the rating may rationalise the choice already made. It
does, however, establish that the mandatory preference is not held in spite of a
belief that voluntary instruments deliver better outcomes.

```{r q13-table}
eff <- full_data |>
  transmute(
    id,
    eff_don  = q13_1,   # voluntary: donations
    eff_cert = q13_2,   # voluntary: market-based certification
    eff_tax  = q13_3,   # mandatory: environmental taxes
    eff_off  = q13_4    # mandatory: mandatory offsetting
  ) |>
  mutate(
    eff_vol     = (eff_don + eff_cert) / 2,
    eff_man     = (eff_tax + eff_off)  / 2,
    # Positive values = respondent judges mandatory instruments the more
    # effective of the two families.
    eff_rel_man = eff_man - eff_vol
  )

eff_means <- tibble(
  Instrument = c("Voluntary donations (Q13\\_1)",
                 "Market certification (Q13\\_2)",
                 "Environmental taxes (Q13\\_3)",
                 "Mandatory offsetting (Q13\\_4)",
                 "Voluntary mean", "Mandatory mean"),
  Mean = sprintf("%.2f", c(
    mean(eff$eff_don,  na.rm = TRUE), mean(eff$eff_cert, na.rm = TRUE),
    mean(eff$eff_tax,  na.rm = TRUE), mean(eff$eff_off,  na.rm = TRUE),
    mean(eff$eff_vol,  na.rm = TRUE), mean(eff$eff_man,  na.rm = TRUE))),
  SD = sprintf("%.2f", c(
    sd(eff$eff_don,  na.rm = TRUE), sd(eff$eff_cert, na.rm = TRUE),
    sd(eff$eff_tax,  na.rm = TRUE), sd(eff$eff_off,  na.rm = TRUE),
    sd(eff$eff_vol,  na.rm = TRUE), sd(eff$eff_man,  na.rm = TRUE)))
)

tt <- t.test(eff$eff_man, eff$eff_vol, paired = TRUE)

eff_vp <- resp_level |>
  left_join(eff, by = "id") |>
  filter(!is.na(eff_rel_man))
m_eff  <- lm(voluntary_pref ~ eff_rel_man, data = eff_vp)
ct_eff <- coeftest(m_eff, vcov = vcovHC(m_eff, "HC3"))

kbl(eff_means, booktabs = TRUE, longtable = FALSE, escape = FALSE,
    align = "lrr",
    caption = paste0(
      "Perceived effectiveness of each financing instrument at delivering ",
      "biodiversity benefits (Q13; 1 = not at all effective, 5 = very ",
      "effective). Mandatory instruments are rated more effective than ",
      "voluntary ones: paired $t = ", sprintf("%.1f", tt$statistic),
      "$, $p < 0.001$, mean difference $= ", sprintf("%.3f", tt$estimate),
      "$. Regressing the VP score on relative perceived mandatory ",
      "effectiveness (mandatory mean minus voluntary mean) gives ",
      "$b = ", sprintf("%.4f", ct_eff["eff_rel_man", "Estimate"]),
      "$ (HC3 SE $= ", sprintf("%.4f", ct_eff["eff_rel_man", "Std. Error"]),
      "$, $p < 0.001$, $n = ", format(nobs(m_eff), big.mark = ","),
      "$): respondents who judge mandatory instruments relatively more ",
      "effective choose voluntary options less often."
    )) |>
  kable_styling(full_width = TRUE, latex_options = "hold_position") |>
  footnote(
    general = "Q13 was administered after the choice tasks, so these associations are post-choice validation and not evidence of mediation.",
    threeparttable = TRUE
  )
```

## S7. Cost vector, income shares, and choke price

The cost attribute spanned SEK 492 to 7,380 per household per year. Table S7
sets those levels against Swedish household disposable income and reports the
bid acceptance curve. Note that acceptance must be measured at the level of the
individual alternative: the share of *tasks* in which either of the two offered
programmes was chosen is mechanically high and is not a bid acceptance curve.

```{r bid-acceptance}
# SCB (Statistics Sweden), table TabVX4bDispInkN, all households, 2024.
SCB_MEDIAN_INCOME_SEK <- 472600
SCB_MEAN_INCOME_SEK   <- 612200

# Alternative-level: was THIS option chosen, at ITS OWN price?
alt <- bind_rows(
  resp |> transmute(chosen = choice_num == 1L, sek = a_price_sek),
  resp |> transmute(chosen = choice_num == 2L, sek = b_price_sek)
)

bid_tab <- alt |>
  group_by(sek) |>
  summarise(n = n(), uptake = mean(chosen), .groups = "drop") |>
  transmute(
    `Cost (SEK/year)`      = format(sek, big.mark = ","),
    `USD/year`             = round(sek / 10),
    `Pct of median income` = sprintf("%.2f", sek / SCB_MEDIAN_INCOME_SEK * 100),
    `Pct of mean income`   = sprintf("%.2f", sek / SCB_MEAN_INCOME_SEK * 100),
    `Alternatives shown`   = format(n, big.mark = ","),
    `Chosen (pct)`         = sprintf("%.1f", uptake * 100)
  )

kbl(bid_tab, booktabs = TRUE, longtable = FALSE, align = "rrrrrr",
    caption = paste0(
      "Cost levels as a share of Swedish household disposable income, with the ",
      "bid acceptance curve. Income reference: median SEK ",
      format(SCB_MEDIAN_INCOME_SEK, big.mark = ","), " and mean SEK ",
      format(SCB_MEAN_INCOME_SEK, big.mark = ","),
      " for all households in 2024 (Statistics Sweden). Uptake is measured per ",
      "alternative, that is the probability a given programme was chosen at its ",
      "own cost, and more than halves across the range."
    )) |>
  kable_styling(full_width = TRUE, latex_options = "hold_position") |>
  footnote(
    general = "The highest cost presented is about 1.6 per cent of median household disposable income.",
    threeparttable = TRUE
  )
```

Willingness to pay is identified where a configuration's **mean
utility-equivalence price**, the annual cost at which mean programme utility
equals mean status-quo utility, falls inside the range presented. This is not a
choke price: at our highest cost 24.2 per cent of alternatives were still
chosen overall, and 48.3 per cent of the most preferred bundle, so demand is
far from choked off. Table S8 reports the threshold for all 24 attribute
combinations. It lies within the bid range for 15 of them. The nine exceptions
all pair public administration monitoring with a mandatory instrument or with
public land. For those bundles the data establish that mean willingness to pay
exceeds SEK 7,380, but the amount above it is extrapolated: cost-vector
experiments show the maximum bid can shift both price sensitivity and estimated
willingness to pay, without a settled direction [@Morkbak2010]. The instrument
ordering is unaffected, being estimated from the utility coefficients rather
than from their ratio to the price coefficient.

The combined non-cost attribute premium of the most preferred bundle is SEK
3,699 (offsetting 913, public land 1,180, public monitoring 1,606), comfortably
inside the span of costs presented (SEK 492 to 7,380). On the face-validity
diagnostic for WTP overshooting proposed by Glenk et al. [@Glenk2024], which compares the
attribute premium of the highest-utility bundle against the cost attribute
range and excludes the opt-out constant, these estimates do not overshoot.

```{r choke-price}
cf_c <- coef(mxl)
asc_c <- unname(cf_c["asc_neither"])
bp_c  <- unname(cf_c["price"])
MAX_BID_SEK <- 7380

grid <- expand_grid(
  instrument = c("Tax", "Donation", "Certification", "Offsetting"),
  land       = c("Small private", "Industrial", "Public"),
  monitoring = c("Private audit", "Public admin")
) |>
  mutate(
    v_src  = c(Tax = 0, Donation = unname(cf_c["don"]),
               Certification = unname(cf_c["cert"]),
               Offsetting = unname(cf_c["off"]))[instrument],
    v_land = c(`Small private` = 0, Industrial = unname(cf_c["ind"]),
               Public = unname(cf_c["pland"]))[land],
    v_mon  = c(`Private audit` = 0, `Public admin` = unname(cf_c["pmon"]))[monitoring],
    # Choke price: cost at which programme utility equals opt-out utility.
    equiv_sek = (asc_c - (v_src + v_land + v_mon)) / bp_c * 1000
  ) |>
  arrange(equiv_sek)

choke_tab <- grid |>
  transmute(
    Instrument = instrument, Land = land, Monitoring = monitoring,
    `Utility-equiv. price (SEK/yr)` = format(round(equiv_sek), big.mark = ","),
    `Pct of median income`  = sprintf("%.2f", equiv_sek / SCB_MEDIAN_INCOME_SEK * 100),
    `Within bid range`      = ifelse(equiv_sek <= MAX_BID_SEK, "yes", "no")
  )

kbl(choke_tab, booktabs = TRUE, longtable = FALSE, align = "lllrrc",
    caption = paste0(
      "Mean utility-equivalence price by attribute configuration, from the ",
      "mixed logit: the annual cost at which mean programme utility equals mean ",
      "status-quo utility. This is not a choke price, since demand is far from ",
      "choked off at that cost. It falls inside the presented range ",
      "(maximum SEK ", format(MAX_BID_SEK, big.mark = ","), ") for ",
      sum(grid$equiv_sek <= MAX_BID_SEK), " of ", nrow(grid),
      " configurations. The exceptions all combine public administration ",
      "monitoring with a mandatory instrument or with public land."
    )) |>
  kable_styling(full_width = TRUE, latex_options = "hold_position",
                font_size = 9) |>
  footnote(
    general = "Estimates for configurations outside the bid range rest on extrapolation beyond the costs any respondent was shown.",
    threeparttable = TRUE
  )
```

## S9. Sensitivity of the cost coefficient to functional form

If the linear-in-cost specification understated price sensitivity at the top of
the range, the utility-equivalence prices in Table S8 would be too high. Table
S9 re-estimates the conditional logit with four treatments of cost: linear,
fully categorical (a dummy per cost level, imposing no shape), piecewise linear
with a knot at SEK 4,920, and quadratic. The categorical specification fits
best, but the improvement is small, and every specification places the most
preferred bundle's utility-equivalence price above the SEK 7,380 maximum bid.

```{r price-spec}
L <- long_db |>
  mutate(
    sek    = price * 1000,                       # long_db holds price in kSEK
    p_hi   = pmax(sek - 4920, 0) / 1000,         # piecewise, knot at SEK 4,920
    price2 = price^2,
    p2460  = as.integer(sek == 2460),
    p4920  = as.integer(sek == 4920),
    p7380  = as.integer(sek == 7380)
  )

base_pars <- c("asc_neither", "don", "cert", "off", "ind", "pland", "pmon")
fit_spec <- function(extra) {
  logitr(data = L, outcome = "chosen", obsID = "obsID",
         pars = c(base_pars, extra), numMultiStarts = 3)
}

specs <- list(
  Linear                       = "price",
  `Categorical (dummy/level)`  = c("p2460", "p4920", "p7380"),
  `Piecewise (knot SEK 4,920)` = c("price", "p_hi"),
  Quadratic                    = c("price", "price2")
)
fits <- lapply(specs, fit_spec)

# Utility-equivalence price for the most preferred bundle. Each specification
# needs its own solution: using the linear coefficient alone would be wrong for
# the quadratic (which has a curvature term) and for the piecewise model when
# the solution lies above the knot, as it does here.
equiv_price <- function(m, form) {
  cf  <- coef(m)
  num <- sum(cf[c("off", "pland", "pmon")]) - unname(cf["asc_neither"])
  if (form == "linear") {
    return(num / (-unname(cf["price"])) * 1000)
  }
  if (form == "piecewise") {
    b1 <- unname(cf["price"]); b2 <- unname(cf["p_hi"]); knot <- 4.92
    p_low <- num / (-b1)                       # solution on the lower segment
    if (p_low <= knot) return(p_low * 1000)
    # num + b1*p + b2*(p - knot) = 0 on the upper segment
    return((-num + b2 * knot) / (b1 + b2) * 1000)
  }
  if (form == "quadratic") {
    a1 <- unname(cf["price"]); a2 <- unname(cf["price2"])
    disc <- a1^2 - 4 * a2 * num                # num + a1*p + a2*p^2 = 0
    if (disc < 0) return(NA_real_)
    roots <- c((-a1 + sqrt(disc)) / (2 * a2), (-a1 - sqrt(disc)) / (2 * a2))
    roots <- sort(roots[roots > 0])
    return(if (length(roots)) roots[1] * 1000 else NA_real_)
  }
  NA_real_                                     # categorical: no cost slope
}
forms <- c("linear", "categorical", "piecewise", "quadratic")

spec_tab <- tibble(
  Specification = names(specs),
  `Log-likelihood` = sprintf("%.1f", sapply(fits, \(m) as.numeric(logLik(m)))),
  k    = sapply(fits, \(m) length(coef(m))),
  AIC  = sprintf("%.1f", sapply(fits, AIC)),
  `AIC vs linear` = sprintf("%+.1f", sapply(fits, AIC) - AIC(fits[[1]])),
  `Utility-equiv. price (SEK)` = mapply(
    \(m, fm) { v <- equiv_price(m, fm)
               if (is.na(v)) "n.a." else format(round(v), big.mark = ",") },
    fits, forms)
)

kbl(spec_tab, booktabs = TRUE, longtable = FALSE, align = "lrrrrr",
    caption = paste0(
      "Sensitivity of the cost coefficient to functional form, conditional ",
      "logit. The categorical specification fits best but by a small margin ",
      "(AIC ", sprintf("%.1f", AIC(fits[[2]]) - AIC(fits[[1]])),
      " relative to linear), and its cost utilities are close to linear in ",
      "cost. In the piecewise model the upper-range slope is almost identical ",
      "to the lower-range slope, so there is little evidence that price ",
      "sensitivity changes across the range presented."
    )) |>
  kable_styling(full_width = TRUE, latex_options = "hold_position",
                font_size = 9) |>
  footnote(
    general = "The utility-equivalence price is not defined for the categorical specification, which estimates no continuous cost slope. All specifications place the most preferred bundle above the SEK 7,380 maximum bid.",
    threeparttable = TRUE
  )
```


# References

::: {#refs}
:::
